---
title: "Reproducible ResearchAssignment 2"
author: "Adlin Fisher Abell Ball"
date: "2023-06-27"
output:
  pdf_document: default
  html_document: default
---
The purpose of this document is to provide an analysis of the data provided
which was taken from the National Oceanic and Atmospheric Administration
and contains information from the 1950 through to November of 2011. In 1996
the NOAA began recording any event types that caused injury to persons or 
damage to property. See the 
[NOAA Website](https://www.ncdc.noaa.gov/stormevents/details.jsp?type=eventtype)
for more information. Then for the sake of consistency of the data collection
process, this analysis will use the data available from the year 1996 and
onward. We will begin by preparing the data for analysis:

# Abstract

The main flaw of this data set is inconsistent labelling. Therefore results 
are contingent on the handling of the event type labels.
Brief analysis of the provided data shows that those events that cause the
most economic damages per occurrence ordered by descending ranking are
hurricanes, drought, and frost. Those events that cause the no event type has
a higher median number of fatalities than any other when considering only
those events that caused either an injury or fatality. Though the events that
cause the greatest total number of casualties, ordered by descending rank, are
tornadoes, floods, and excessive heat, excessive heat has the highest median
number of injuries at 10, while fog and winter storms are tied for the next
highest medians of 4 injuries.
  
# Data Processing
  
The first step is naturally to load in the data and take the section that we
want for analysis. 
```{r}
library(ggplot2)
library(tidyverse)
damage = read.csv(bzfile("repdata_data_StormData.csv.bz2"))
damage$BGN_DATE = strptime(damage$BGN_DATE, tz='', '%m/%d/%Y %H:%M:%OS')
damage96 = damage[damage$BGN_DATE > strptime('1996-01-01', '%F'),
                  1:length(colnames(damage))]
remove(damage)
```

Now, to begin we must prepare the data for calculating the financial impact.
For this, I drop all the irrelevant rows in order to save time in operating on
the data. Note that the `PROPDMGEXP` and `CROPDMGEXP` columns contain a factor
that encodes the amount of damage in the `PROPDMG` and `CROPDMG` columns 
respectively. In this scheme, `K` indicates that the column damages are in
thousands of dollars, `M` indicates millions, and `B` indicates billions. There
are other values in the encoding columns, but since these are not accounted for
in the
[data documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf),
the rows associated with these entries are discarded.

```{r}
# First, for efficiency's sake I select only the relevant columns for this 
# analysis.
columns = c('BGN_DATE', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP',
            'EVTYPE', 'REFNUM')
QuantDam = damage96[columns]

#Next, I will discard all of the data that does not have its costs properly
# prepared.
QuantDam = QuantDam[(QuantDam$PROPDMG == 0)|
                      (QuantDam$PROPDMGEXP %in% c('K', 'M', 'B', '')), 
                    1:length(colnames(QuantDam)) ]
QuantDam = QuantDam[(QuantDam$CROPDMG == 0)|
                      (QuantDam$CROPDMGEXP %in% c('K', 'M', 'B', '')), 
                    1:length(colnames(QuantDam)) ]

totaldamage = rep(0, dim(QuantDam)[1])
ttldmg2023 = rep(0, dim(QuantDam)[1])

# Below I combine the total damages to both crops and property while calculating
# the costs.

pks = (QuantDam$PROPDMGEXP == 'K' | QuantDam$PROPDMGEXP == 'k')
totaldamage = totaldamage + QuantDam$PROPDMG * pks * 10^3
remove(pks)

pms = (QuantDam$PROPDMGEXP == 'M')
totaldamage = totaldamage + QuantDam$PROPDMG * pms * 10^6
remove(pms)

pbs = (QuantDam$PROPDMGEXP == 'B')
totaldamage = totaldamage + QuantDam$PROPDMG * pbs * 10^9
remove(pbs)

cks = (QuantDam$CROPDMGEXP == 'K' | QuantDam$CROPDMGEXP == 'k')
totaldamage = totaldamage + QuantDam$CROPDMG * cks * 10^3
remove(cks)

cms = (QuantDam$CROPDMGEXP == 'M')
totaldamage = totaldamage + QuantDam$CROPDMG * cms * 10^6
remove(cms)

cbs = (QuantDam$CROPDMGEXP == 'B')
totaldamage = totaldamage + QuantDam$CROPDMG * cbs * 10^9
remove(cbs)
```

Since the value of a dollar is not constant accross time, I also wish to 
account for inflation in my analysis of economic damages. To do this, I used
the
[Bureau of Labor Statistics CPI Database](https://beta.bls.gov/dataViewer/view/timeseries/CUSR0000SA0) 
(accessed June 12, 2023) to get the inflation index. The calculations to
convert the damages to May 2023 dollars are below:

```{r}
# First, I initialize a couple vectors for our calculations.
eventCPI = rep(0, dim(QuantDam)[1])
ttldmg2023 = rep(0, dim(QuantDam)[1])

# Now I read in the CPI data.
QuantDam = cbind(QuantDam, TOTALDMG = totaldamage)
CPI = read.csv('CPI.csv')
```

Now I convert the times to date-time objects. Then I take the weather event
beginning dates and convert them to the first of their month in order to match
the CPI for that date to the data from the CPI database. Then I remove
superfluous vectors since they are a non-trivial amount of storage.

```{r}
CPI$Label = strptime(paste(CPI$Label, '01'), '%Y %b %d')
CPIperiod = strptime(paste(format(QuantDam$BGN_DATE, "%Y-%m"),
                           '01', sep= '-'), format = '%Y-%m-%d')
periodind = match(CPIperiod, CPI$Label)
for (i in 1:length(CPIperiod)) {
  eventCPI[i] = CPI$Value[periodind[i]]
}
remove(CPIperiod, periodind)

# Lastly, take the most current CPI at time of writing and calculate the total
# total damages for each event in May 2023 Dollars.
NowCPI = CPI$Value[match(strptime('2023-05-01', '%Y-%m-%d'), CPI$Label)]
ttldmg2023 = (NowCPI/eventCPI) * totaldamage
QuantDam = cbind(QuantDam, TOTAL2023DMG = as.numeric(ttldmg2023))

remove(totaldamage, CPI, NowCPI)
```

Next, since the primary purpose of this analysis is to evaluate events by their
type, (`EVTYPE` in the data itself), we need to address some problems with this
column. Firstly, the events appear to be inconsistently labelled, with
capitalization and spacing varying inconsistently across labels, causing R to
not recognize them as the same type of event. Secondly, there appear to be
tags that are used either in the case of having one type of event *or* another,
which are also used as labels for events that are *combinations* of those
events. By reviewing the events with the highest amount of economic damages,
I've selected some event tags below to relabel to broader event tags in which
they fit.

```{r}
# By using the `grep` function, I search each event label for occurrences of the
# first argument and change the label to a more inclusive one.

QuantDam$EVTYPE[grep('hurricane', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = 'HURRICANE/TYPHOON'
QuantDam$EVTYPE[grep('typhoon', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = 'HURRICANE/TYPHOON'
QuantDam$EVTYPE[grep('damaging freeze', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = 'DAMAGING FREEZE'
QuantDam$EVTYPE[grep('flood', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = 'FLOOD'
QuantDam$EVTYPE[grep('fld', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = 'FLOOD'
QuantDam$EVTYPE[grep('thunderstorm', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = "THUNDERSTORM"
QuantDam$EVTYPE[grep('tstm', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = "THUNDERSTORM"
QuantDam$EVTYPE[grep('freeze', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = "FROST/FREEZE"
QuantDam$EVTYPE[grep('frost', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = "FROST/FREEZE"
QuantDam$EVTYPE[grep('storm surge', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = 'STORM SURGE/TIDE'
QuantDam$EVTYPE[grep('high tide', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = 'STORM SURGE/TIDE'
QuantDam$EVTYPE[grep('wild fire', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = 'WILD/FOREST FIRE'
QuantDam$EVTYPE[grep('wildfire', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = 'WILD/FOREST FIRE'
coldtags = unique(QuantDam$EVTYPE)[grep('cold', 
                                        unique(QuantDam$EVTYPE),
                                        ignore.case = TRUE)]
coldtags
```

As we can see above, there are multiple redundant tags for cold whether or
windchill, but not every event descriptor that contains the word "cold" is 
equivalent. Since all of these event titles except "COLD AND SNOW" appear to be
used to describe events that are characterized by cold tempuratures, windchill,
or some combination of the two, it is intuitive to combine them. Note that since
cold and snow includes snow in its characteristic conditions rather than simply
cold temperatures, it will be excluded.

```{r}
coldtags = coldtags[coldtags != 'COLD AND SNOW']

for (i in 1:length(coldtags)) {
  QuantDam$EVTYPE[grep(coldtags[i], QuantDam$EVTYPE,
                     ignore.case = TRUE)] = "COLD/WIND CHILL"
}
QuantDam$EVTYPE[grep('windchill', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = "COLD/WIND CHILL"

QuantDam$EVTYPE[grep('wind chill', QuantDam$EVTYPE,
                     ignore.case = TRUE)] = "COLD/WIND CHILL"

# Remove entries that missused the EVTYPE entry.
QuantDam = QuantDam[grep('summary', QuantDam$EVTYPE, ignore.case = TRUE,
                         invert = TRUE),]

# Now I consolidate the damages by the aggregate function to select the top 15
# 

totaldamages = aggregate(QuantDam$TOTAL2023DMG, list(QuantDam$EVTYPE),
                         FUN = sum)
names(totaldamages) = c('evtype', 'dmg')
top15 = order(totaldamages$dmg, decreasing = TRUE)[1:15]
QuantDam1 = QuantDam[QuantDam$EVTYPE %in% totaldamages$evtype[top15] &
                       QuantDam$TOTALDMG != 0, ]

#Below I calculate some summary statistics for use in the final analysis.

hurmed = median(QuantDam1$TOTAL2023DMG[QuantDam1$EVTYPE == 'HURRICANE/TYPHOON'])
drtmed = median(QuantDam1$TOTAL2023DMG[QuantDam1$EVTYPE == 'DROUGHT'])
fstmed = median(QuantDam1$TOTAL2023DMG[QuantDam1$EVTYPE == 'FROST/FREEZE'])
napadmg = format(max(QuantDam1$TOTAL2023DMG[QuantDam1$EVTYPE == 'FLOOD']),
                 big.mark = ',', scientific = FALSE)
fldmed = format(median(QuantDam1$TOTAL2023DMG[QuantDam1$EVTYPE == 'FLOOD']),
                big.mark = ',')

# Tossing objects that are no longer needed
remove(eventCPI, totaldamages, QuantDam)
```

Now for the human damage of these events. For this we will once again drop the
irrelevant categories. For consistency's sake, we will consolidate the event
type tags the same way as we did previously. Not surprisingly, the events that
cause the most economic damages have a large amount of overlap with those that
caused the most casualties. Note that the table for casualties drops the total
casualties column and reformats the data below for plotting.

```{r}
HumDmg = damage96[, c('REFNUM', 'EVTYPE', 'INJURIES', 'FATALITIES')]
HumDmg$EVTYPE[grep('hurricane', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = 'HURRICANE/TYPHOON'
HumDmg$EVTYPE[grep('typhoon', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = 'HURRICANE/TYPHOON'
HumDmg$EVTYPE[grep('damaging freeze', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = 'DAMAGING FREEZE'
HumDmg$EVTYPE[grep('flood', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = 'FLOOD'
HumDmg$EVTYPE[grep('fld', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = 'FLOOD'
HumDmg$EVTYPE[grep('thunderstorm', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = "THUNDERSTORM"
HumDmg$EVTYPE[grep('tstm', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = "THUNDERSTORM"
HumDmg$EVTYPE[grep('freeze', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = "FROST/FREEZE"
HumDmg$EVTYPE[grep('frost', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = "FROST/FREEZE"
HumDmg$EVTYPE[grep('storm surge', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = 'STORM SURGE/TIDE'
HumDmg$EVTYPE[grep('high tide', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = 'STORM SURGE/TIDE'
HumDmg$EVTYPE[grep('wild fire', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = 'WILD/FOREST FIRE'
HumDmg$EVTYPE[grep('wildfire', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = 'WILD/FOREST FIRE'
coldtags = unique(HumDmg$EVTYPE)[grep('cold', 
                                        unique(HumDmg$EVTYPE),
                                        ignore.case = TRUE)]
coldtags = coldtags[coldtags != 'COLD AND SNOW']

for (i in 1:length(coldtags)) {
  HumDmg$EVTYPE[grep(coldtags[i], HumDmg$EVTYPE,
                     ignore.case = TRUE)] = "COLD/WIND CHILL"
}
HumDmg$EVTYPE[grep('windchill', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = "COLD/WIND CHILL"

HumDmg$EVTYPE[grep('wind chill', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = "COLD/WIND CHILL"
HumDmg$EVTYPE[grep('rip currents', HumDmg$EVTYPE,
                     ignore.case = TRUE)] = "RIP CURRENT"
HumDmg$EVTYPE[HumDmg$EVTYPE != "COLD/WIND CHILL"][grep('wind', 
                   HumDmg$EVTYPE[HumDmg$EVTYPE != "COLD/WIND CHILL"], 
                   ignore.case = TRUE)]= 'WIND'
HumDmg$EVTYPE[HumDmg$EVTYPE == 'WND' ]= 'WIND'
HumDmg$EVTYPE[grep('record temperature', HumDmg$EVTYPE,
                   ignore.case = TRUE)] = 'EXCESSIVE HEAT'
HumDmg$EVTYPE[grep('record temperature', HumDmg$EVTYPE,
                   ignore.case = TRUE)] = 'EXCESSIVE HEAT'
HumDmg$EVTYPE[grep('heat', HumDmg$EVTYPE,
                   ignore.case = TRUE)] = 'EXCESSIVE HEAT'
NotWet = grep('wet', HumDmg$EVTYPE, ignore.case = TRUE, invert = TRUE)

HumDmg$EVTYPE[NotWet][grep('warm', HumDmg$EVTYPE[NotWet],
                           ignore.case = TRUE)] = 'EXCESSIVE HEAT'
remove(NotWet)

# Remove entries that missused the EVTYPE entry.
HumDmg = HumDmg[grep('summary', HumDmg$EVTYPE, ignore.case = TRUE,
                     invert = TRUE),]

CASUALTIES = HumDmg$INJURIES + HumDmg$FATALITIES

HumDmg = cbind(HumDmg, CASUALTIES)

totalcasualties = aggregate(HumDmg[,c('INJURIES', 'FATALITIES', 'CASUALTIES')], 
                            list(HumDmg$EVTYPE), sum)

#holding on for troubleshooting
#totalcasualties$EVTYPE[order(totalcasualties$CASUALTIES, decreasing = TRUE)]

top15 = order(totalcasualties$CASUALTIES, decreasing = TRUE)[1:15]
colnames(totalcasualties) = c('EVTYPE', 'INJURIES', 'FATALITIES', 'CASUALTIES')
top15casualties = totalcasualties$EVTYPE[top15]

HumDmg = HumDmg[HumDmg$CASUALTIES != 0,c('EVTYPE', 'INJURIES', 'FATALITIES')]
graphhmdmg = pivot_longer(HumDmg, c('FATALITIES', 'INJURIES'))
graphhmdmg = graphhmdmg[graphhmdmg$EVTYPE %in% top15casualties,]
graphhmdmg = graphhmdmg[graphhmdmg$value != 0,]
```

## Results

```{r}
ggplot(data = QuantDam1) +
  geom_boxplot(mapping = aes(x = reorder(EVTYPE , TOTAL2023DMG+1, median),
                             TOTAL2023DMG, fill = EVTYPE,
                             color = EVTYPE),
               outlier.color = 'dark gray') +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
   scale_fill_manual(values = rep("light blue", 15)) +
   scale_color_manual(values = rep('black',15)) +
  scale_y_log10(name = "Median Damage \n (log+1 Dollars as of May, 2023)") +
  scale_x_discrete(name = 'Event Type Ordered by Median Damage')
```

Note that the above graph is of only the events that did some amount of damage,
with events that did no damage excluded. With that in mind, we can see that the 
event category that has the highest median damages is hurricanes and typhoons 
with a median of $`r format(hurmed, big.mark=',')`.
The categories with the second and third highest median damages are drought,
with median damages of $`r format(drtmed, digits = 2, big.mark=',')`, and 
freezing/frosts, with a median of $`r format(fstmed, digits = 2, big.mark=',')`.
This graph also reveals a number of outlying points. The most prominent is the
event with the single highest damage, which comes from the flooding category.
This event's damages total to $`r napadmg`, the highest of any single event.
Compare this with the flood category's median of $`r fldmed`. This event is the
Napa Valley flooding of 2005.

```{r}
ggplot(data = graphhmdmg[graphhmdmg$value != 0,]) +
   geom_jitter(mapping = aes(x = reorder(EVTYPE, value, sum),
               y = value), size = .05, color = '#3545db88') +
  geom_boxplot(mapping = aes(x = reorder(EVTYPE , value, sum),
               y = value), color = '#b3140588', fill = '#f0403088',
               outlier.colour = '#b3140544') + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
   scale_fill_manual(values = rep(c("light blue", 'red'), 15)) +
   scale_color_manual(values = rep('black',30)) +
  scale_y_log10(name = "Casualties by Type \n (log+1 Number of Casualties)") +
  scale_x_discrete(name = 'Event Type Ordered by Total Casualties') +
  facet_grid(name ~ ., scales = 'free_y')
```

Due to the fact that boxplots alone don't appear to give a full view of these
data, I plotted both a box plot and the individual points on the above graph.
Once again, I've omitted events where no damage was recorded. 
We can see above that while tornadoes, flood, and abnormal heat lead all other
event types in terms of total casualties, they are not as a rule the events
with the highest mean damages by either category. While tornadoes have the
highest combined casualties of any event, the median number of injuries for 
that category is 3, compared with a median of 10 injuries in the category of 
excessive heat events. Note also that both fog and winter storm events have
higher median injuries at 4 as well. The medians in the category of fatalities 
are notable in that they are all the same, with a median of 1. Therefore,
excessive heat is one of the most hazardous weather event types for human
health, along with floods and tornadoes. While these events don't consistently 
do damage more than other event types, the amount of damage that they do incur
when they do have impacts on humans total to a higher amount than any other
events.